Reproducible Research: Peer Assessment 2
------------------------------------------


## Impact of Severe Weather Events on Public Health and Economy in the United States 
 
### Synonpsis
In this report, we try to analyze the impact of different weather events that have a greater impact on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011.This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences. 
 
## Visible settings  
```{r}  
echo = TRUE 
options(scipen = 1) 
```

## Loading libraries
```{r}
library(R.utils) 
library(ggplot2) 
library(plyr) 
library(gridExtra) 
``` 

## Data Processing
```{r}
if (!"repdata_data_StormData.csv.bz2" %in% dir("/dataset/")) 
{ 
     bunzip2("dataset/repdata_data_StormData.csv.bz2", overwrite=T, remove=F) 
} 
```

### Reading Data
```{r} 
stormData <- read.csv("dataset/repdata_data_StormData.csv", sep = ",")
dim(stormData)
```

There are 902297 rows and 37 columns total in the database. The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

```{r} 
if (dim(stormData)[2] == 37) 
{ 
   stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y")) 
} 
hist(stormData$year, breaks = 30)
```  

Based on the above histogram, we see that the number of events tracked starts to significantly increase around 1995. So, we use the subset of the data from 1990 to 2011 to get most out of good records. 
```{r} 
storm <- stormData[stormData$year >= 1995, ] 
dim(storm) 
``` 

#### Impact on Public Health

In this section, we check the number of **fatalities** and **injuries** that are caused by top twenty severe weather events. 
```{r} 
sortData <- function(fieldName, top = 20, dataset = stormData) 
  { 
     index <- which(colnames(dataset) == fieldName) 
     field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum") 
     names(field) <- c("EVTYPE", fieldName)
     field <- arrange(field, field[, 2], decreasing = T)
     field <- head(field, n = top) 
     field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE)) 
     return(field) 
  } 

fatalities <- sortData("FATALITIES", dataset = storm)
injuries <- sortData("INJURIES", dataset = storm)
```
#### Impact on Economy 
We will convert the **property damage** and **crop damage** data into comparable numerical forms according to the meaning of units described in the code book. Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observations. 

```{r} 
convertData <- function(dataset = storm, fieldName, newFieldName) 
{ 
     totalLen <- dim(dataset)[2] 
     index <- which(colnames(dataset) == fieldName) 
     dataset[, index] <- as.character(dataset[, index]) 
     logic <- !is.na(toupper(dataset[, index])) 
     dataset[logic & toupper(dataset[, index]) == "B", index] <- "9" 
     dataset[logic & toupper(dataset[, index]) == "M", index] <- "6" 
     dataset[logic & toupper(dataset[, index]) == "K", index] <- "3" 
     dataset[logic & toupper(dataset[, index]) == "H", index] <- "2" 
     dataset[logic & toupper(dataset[, index]) == "", index] <- "0" 
     dataset[, index] <- as.numeric(dataset[, index]) 
     dataset[is.na(dataset[, index]), index] <- 0 
     dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index]) 
     names(dataset)[totalLen + 1] <- newFieldName 
     return(dataset) 
 }  
 
storm <- convertData(storm, "PROPDMGEXP", "propertyDamage") 
storm <- convertData(storm, "CROPDMGEXP", "cropDamage") 
options(scipen=999) 
property <- sortData("propertyDamage", dataset = storm) 
crop <- sortData("cropDamage", dataset = storm) 
```

### Results 
As for the impact on public health, we have got two sorted lists of severe weather events below by the number of people badly affected. 
```{r} 
fatalities 
injuries 
``` 

Plotting graphs for total fatalities and total injuries affected by these severe weather events.  

### Fatalities Plot
```{r} 
qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar", binwidth = 1) +
   scale_y_continuous("Number of Fatalities") +  
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
   xlab("Severe Weather Type") +  
   ggtitle("Total Fatalities by Severe Weather Events \n in the U.S. from 1995 - 2011") 

```
### Injuries Plot
```{r}
qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar", binwidth = 1) +  
   scale_y_continuous("Number of Injuries") +  
   theme(axis.text.x = element_text(angle = 45,  
   hjust = 1)) + xlab("Severe Weather Type") +  
   ggtitle("Total Injuries by Severe Weather Events \n in the U.S. from 1995 - 2011") 
```

Based on the above histograms, we found that **excessive heat** and **tornado** cause most fatalities; **tornato** causes most injuries in the United States from 1995 to 2011. 
 
As for the impact on economy, we got two sorted lists below by the amount of money cost by damages.
 
```{r} 
property 
crop 
``` 

Plotting graphs for total property damage and total crop damage affected by these severe weather events.

### Property Plot
```{r} 
qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar", binwidth = 1) +  
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+  
   xlab("Severe Weather Type") + ggtitle("Total Property Damage by Severe Weather Events in\n the U.S. from 1995 - 2011") 
```

### Crop Plot
```{r}
qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar", binwidth = 1) +  
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") +  
   xlab("Severe Weather Type") + ggtitle("Total Crop Damage by Severe Weather Events in\n the U.S. from 1995 - 2011") 
```

Based on the above histograms, we found that **flood** and **hurricane/typhoon** cause most property damage; **drought** and **flood** causes most crop damage in the United States from 1995 to 2011. 
 
### Conclusion   
From these data, we found that **excessive heat** and **tornado** are most harmful with respect to population health, while **flood**, **drought**, and **hurricane/typhoon** have the greatest economic consequences.